### 2023-04-13 submission/manuscript.Rmd
Editing through Introduction

```{r, echo=FALSE}
options(tidyverse.quiet = TRUE)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggtext))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(glue))

@@

# Waste not, want not: Revisiting the analysis that called rarefaction into question

@@

Since the development of sequencing technologies such as those provided by 454 and Illumina, microbiome researchers have struggled to produce a consistent number of sequences from each sample in a dataset. It is common to observe more than 10-fold variation in the number of sequences per sample [XXXXX]. Regardless of the source of this variation, researchers desire approaches to control for uneven sampling effort. Of course, this desire is not unique to microbiome research and is a challenge faced by all community ecologists. Common approaches to controlling uneven sampling efforts have included use of proportional abundance (i.e., relative abundance), normalization of counts, parameter estimation, and rarefaction.

In 2014 Paul McMurdie and Susan Holmes published their "Wast not, want not: why rarefying microbiome data is inadmissible" (WNWN) in PLOS Computational Biology [XXXXX]. This paper has had a significant impact on the approaches that microbiome researchers use to analyze 16S rRNA gene sequence data. According to Google Scholar, this paper has been cited more than 2,300 times as of January 2023. Anecdotely, I have received correspondence from researchers over the past 10 years asking how to address critiques from reviewers who criticize my correspondents' analysis for rarefying (e.g., see [this Twitter thread](https://twitter.com/inanna_nalytica/status/1264679859672006656)). I have also received these types of comments from reviewers, specifically in regards to a preprint that I posted in 20XX in regards to my critique of the practice of removing rare taxa from analyses [XXXXX]. In the process of responding to these critiques and preparing a manuscript investigating rarefaction and other approaches to control for uneven sequencing effort, I decided to reassess the WNWN study including their definitions, simulations, and analyses.

@@

As I attempted to reproduce the results of WNWN, I noticed that the step that purported to rarefy the data only performed one subsampling of the data (Lines 404 through 416 of `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd`). This caused me to re-inspect how McMurdie and Holmes defined "rarefying" in the following quoted text from their paper:

@@

It was unfortunate that McMurdie and Holmes used the term "rarefying" here and throughout their manuscript. The authors were correct to state that the distinction between "rarefying" and "rarefaction" is confusing and leads to their conflation. In my experience, subsequent researchers have conflated the results of this study of the effects of rarefying data with rarefaction of data. As an example, Willis (XXXXX) describes problems with rarefaction rather than rarefying data when citing WNWN in her paper proposing alternatives to rarefaction for use with alpha diversity data:

@@

Adding to the confusion is that the papers cited in the first sentence of the quote I  WNWN included above either do not use the words "rarefy" or "rarefying" or use them interchangably with "rarefaction". In hindsight, as shown in the quoted text, McMurdie and Holmes do emphasize the distinction between rarefying and rarefaction. However, because they seem to have coined a new meaning for rarefying, they seem to have only added to the confusion by using the generally used verb form of rarefaction. Further confusion comes from the author's admonition in the final sentence that some researchers have failed to repeat the subsampling step. To most scientists, repeating the subsampling step is rarefaction. My preference is to use subsampling as the term describing the process they refer to as rarefying. In other words rarefaction with a single randomization.

To provide a more clear definition of rarefaction, I propose the following:

@@

This definition aligns well with how rarefaction was originally defined for comparing richness (i.e., the number of taxa in a community) across communities when communities are sampled to different depths. It is important to note that this procedure generates substantially different results to those obtained without accounting for uneven sampling effort or using relative abundances, normalized counts, compositional data transformatins, variance stabilization procedures, or estimation techniques. I have explored the differences in results obtained using the diversity of approaches for controlling for uneven sampling effort; rarefaction, as described here, outperforms the other approaches [XXXXX].

With this more general approach to rarefaction, rarefaction can be performed using any alpha or beta diversity metric. This strategy has been widely used by my research group and others. The procedure outlined above could also be used for hypothesis tests of differential abundance; however, thought would need to be given to how to synthesize the results of these tests across a large number of replications.

@@

McMurdie and Holmes analyzed the effect of rarefying and other approaches on clustering accuracy using what they called "Simulation A" in their Figure 2A and elsewhere in their paper. In Simulation A, they investigated the ability to correctly assign samples to one of two clusters using simulated data used to generate 40 samples from each of two distributions. A variety of approaches were used to calculate distances between the samples those distances were used as input to partitioning around mediods (PAM) clustering with two groups. The accuracy of the cluster assignment was used as the metric to assess performance. This analysis was performed in the `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` R-flavored markdown file that was published as Protocol S1 in the original paper. Line numbers from this file will be refernced with an "L" as a prefix.

The two distributions were generated using human fecal and ocean data originally take from the GlobalPaterns dataset (L129) [XXXXX]. To generate a fecal and ocean template distribution, the authors included any operational taxonomic unit (OTU) that appeared in more than one of the 4 fecal and 3 ocean samples (L60 and L137). The OTUs were sorted by how many of the 7 samples the OTUs were observed in followed by their total abundance across all 7 samples (L139). From this sorted list they identified the identifiers of the first 2000 OTUs (L66). Returning to the 7 samples they selected the 2000 most common and abundant OTUs and pooled the abundances of the fecal and ocean samples separately to create two templates (L144, L159-160, L197-198).

Next, the fecal and ocean templates were mixed in 8 different fractions to generate two community types that differend by varying effect sizes (1, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5; L170-195, L220). To simulate the variation in sequencing depth across the 80 samples, they normalized he number of sequences from each of the `gp_n_samples` samples in the GlobalPatterns dataset so that the median number of sequences (N~L~) for the GlobalPatterns had 1,000, 2,000, 5,000, or 10,000 sequences (L324-325). They then randomly sampled the `gp_n_samples` normalized sequencing depths to generate 80 sampling depths. From each community type, they simulated 40 samples by sampling to the desired number of reads (L73, L230-233 and L326-327). Each simulation condition was repeated 5 times (L85). This resulted in 160 simulations (8 effect sizes x 4 median sampling depths x 5 replicates = 160 simulations). Finally, they removed rare and low prevalence OTUs in two steps. First, they removed any OTUs whose total abundance was less than 3 across all 80 samples and that did not appear in at least 3 samples (L368-386). Second, they removed any OTUs that did not have more than 1 sequence in more than 5% of the 80 samples (i.e., 4 samples) and that did not have a total abundance across the 80 samples greater than one half of the number of samples in each community type (i.e., 20) (L523-538, L551).

@@

1. Simulated conditions were only replicated 5 times each, which caused results to be sensitive to random number generator seed
2. The average sizes of the libraries were small by modern standards
3. DeSeq-based variance stabilization was used with inappropriate distance calculation methods
4. A single subsampling of each dataset was evaluated rather than using rarefaction, which likely resulted in noisier data
5. Results using PAM clustering were not directly compared to those of K-means and hierarchical clustering
6. Subsampling removed the smallest 15% of the samples, which penalized accuracy values by 15 percentage points
7. The distribution of library sizes was poorly chosen
8. A filtering step was applied to remove rare taxa from the simulated datasets, which could distort the shape of the communities
9. No test of effect of transformation methods to account for effects of uneven sample size between treatment groups
10. Clustering accuracy was used rather than direct comparisons beta diversity
11. No consideration of effects of transformations on alpha diversity metrics

@@

Each simulated condition was replicated 5 times in WNWN and the paper reports the mean and standard deviation of the replicate clustering accuracies. The relatively small number of replicates accounts for the jerkiness of the lines in the original Figures 4 and 5 (e.g. the Bray-Curtis distances calculated on the DEeqVS transformed data). A better approach would have been to use 100 replicates as this would reduce the dependency of the results on the random number generator's seed. By increasing the number of replicates it was also possible to compare the probability of falsely and correctly clustering samples from the same and different treatment groups together (see points 9-11, below). Because the accuracies are unlikely to be symmetric around a mean at larger accuracy values the median and 95% confidence intervals or intraquartile range should have been reported. To test the effect of increasing the number of replicates, I pulled apart the code in `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` into individual R and bash scripts that were executed using a Snakemake workflow with the same Conda environment as above. This was necessary since the number of simulated conditions because of this change increased 20-fold. Such intense data processing was not practical within a single R-flavored markdown document. Again, the observed results were qualitiatively similar to those generated using the single R-flavored markdown file (Figures S4 (**pam_subsample15_fig_4.pdf**) and S5(**pam_subsample_fig_5.pdf**) ). The increased number of replications resulted in smoother lines and allowed me to present empirical 95% confidence intervals. For all analyses in the remainder of this paper, I used 100 randomized replicates per condition.

@@

### 3. DeSeq-based variance stabilization was used with inappropriate distance calculation methods

Close comparison of the original Figure 4 and my version (**pam_subsample15_fig_4.pdf**) revealed one important difference between the two plots. In the original analysis, the accuracies for the Weighted UniFrac distances at the largest effect size (i.e., 3.5) were 1.0 for median sequencing depths of `r oxford_comma(deseq_wunifrac$n_seqs)`. In my version of the analysis, the values for the same sequencing depths were `r oxford_comma(deseq_wunifrac$accuracy)`, respectively. The 95% confidence interval for these accuracies spaned `r min(deseq_wunifrac$lci)` and `r max(deseq_wunifrac$uci)`. The Bray-Curtis distances were also different by both methods at smaller effect sizes and had wide confidence intervals. Inspection of the DeSeq normalized OTU counts revealed that the method resulted in negative values. In fact, rendering the R-flavored markdown files in WNWN's Protocol S1 generated a warning message when passing the DeSeq normalized counts to the Bray-Curtis calculator, which said, "results may be meaningless because data have negative entries in method 'bray'". Although the weighted UniFrac calculator function did not generate a similar warning message, negative values would also result in similarly meaningless distances. Both are due to the fact that the calculators sum the counts of each OTU in both samples being compared. In contrast, a Euclidean distance does not use a similar sum, but sums the square of the difference between the OTU abundance in each sample.

To assess the prevalence of negative counts in the simulated data, I quantified the fraction of negative values in the OTU matrix from each simulation and counted the number of simulations where the transformed OTU table had at least one negative value (Figure **deseq_negative_value.pdf**). In general the fraction of negative OTU counts increased with effect size, but decreased with sequencing effort. The fraction of simulations with at least onee negative value increased with effect size and sequencing effort. The high frequency of negative OTU counts resulted in highly variable Bray-Curtis and weighted UniFrac values. It is likely that because the WNWN analysis only used 5 replicates that the large variation in accuracies at high effect sizes was missed initially. For the rest of this reanalysis study, I will only report results using the DeSeq-based variance stabilization transformation with the Euclidean distance.

### 4. A single subsampling of each dataset was evaluated rather than using rarefaction

@@

### 6. Subsampling removed the smallest 15% of the samples

In WNWN, the authors quantified the tradeoff between sampling effort, the number of samples removed below the threshold, and clustering accuracy (original Figure 5, my Figure S5). Although the optimal sampling effort varied by distance metric, transformation method, and sampling effort, they removed samples whose number of sequences was less than the 15th percentile (L404-419). They acknowledged that this screening step, which was only used with subsampling, would decrease clustering accuracy putting it at a relative disadvantage to the other methods (page 5, column 1, last paragraph). Therefore, it was not surprising that the peak clustering accuracy for their subsampled data was at 85% (Figure S2). Because the true result would not known *a priori* in microbiome studies, it would be impossible for researchers to conduct a sensitivity analysis comparing the tradeoffs between sequencing depth, sample number, and clustering accuracy to select a sampling depth. [**Although the authors claim that "Rarefying counts requires an arbitrary selection of a library size minimum that affects downstream inference" (page 8, column 1, point 3), in acutal microbiome studies the selection of a sampling depth is not as arbitrary as the authors claim. Rather, to avoid p-hacking, researchers pick a set of criteria where they will include or exclude samples prior to testing their data.**] The differences in clustering accuracy between subsampling and rarefaction and using PAM and K-means clustering indicated that it was necessary to reassess the tradeoff between the library size minimum quantile and clustering accuracy. When using rarefaction, K-means clustering, and only considering conditions with 2,000 or more sequences, there was not a condition where setting a higher threshold resulted in a better accuracy than using all of the samples (**kmeans_rarefaction_fig_5.pdf**). These results showed that for modern sequencing depths, using the full datasets with rarefaction and K-means clustering resulted in accuracies that were typically better than those observed when removing the smallest 15% of the samples from each simulated dataset. When the original Figure 4 was recast with these approaches, Rarefaction performed at least as well as any of the other tranformations with each distance calculation, except for with the Poisson distance (Figure **kmeans_rarefaction15_fig_4.pdf**). It is worth noting that at larger effect sizes, K-means clustering did not perform as well for some combinations of normalization methods and distance calculations (compare **kmeans_rarefaction15_fig_4.pdf** and **pam_subsample00_fig_4.pdf**); however, those combinations that performed worse by K-means were not as good as rarefaction or subsampling by either clustering method.

@@

### 7. The distribution of library sizes was poorly chosen

As described above, the sequencing depths used in the `r gp_n_samples` GlobalPatterns datasets were used as the distribution to create sequencing depths for the 80 samples that were generated in each simulation. The GlobalPatterns datasets had a mean of `r gp_mean` sequences and a median of `r gp_median` sequences per dataset. The datasets ranged in sequencing depth between `r gp_min` and `r gp_max` sequences for a `r gp_max/gp_min`-fold difference. Rather than representing a typically observed distribution of sequencing depths that would be skewed right, the sampling distribution was normally distributed (Shapiro-Wilk test of normality, P=`r shapiro.test(gp_counts$n_seqs)$p.value`) (Figure ***distribution_shape.pdf***). From these simulations it is unclear how sensitive the various transformations and distance calculations are to a skewed distribution. A second limitation of this sampling distribution is that it only contained `r gp_n_samples` unique sampling depths such that each sampling depth would have been re-used an average of `r 80/gp_n_samples` times in each simulation. Yet, it is unlikely for a real sequence collection to have duplicate sequencing depths.

I created a new set of simulations that would focus on solely on the effect of the shape of the distribution on the results to reassess the WNWN results in the context of a more typical distribution of sample sizes. I created a simple sequencing depth distribution where there were 80 depths logarithmically distributed between the minimum and maximum sequencing depths of the GlobalPatterns dataset (Figure ***distribution_shape.pdf***). The median of this distribution was `r log_median` and the mean was `r log_mean`. When I regenerated the original Figures 4 and 5 using the log-distributed sequencing effort distribution, the differences in transformation methods were more apparent (Figure ***fig_4_kmeans_rarefaction00_log.pdf***). For each of the distance calculators, rarefying the data to the size of the smallest dataset yielded accuracies that were at least as good as the other methods across effect sizes and median sequencing depths. The difference was most pronounced at smaller effect sizes and sequencing depths. When comparing the performance of rarefaction across distance calculators for different effect sizes, sequencing depths and size of smallest sample (***fig_5_kmeans_rarefaction_log.pdf***), the accuracies I observed using the log-distributed sample sizes was at least as good as those obtained using the GlobalPatterns-based distribution ***fig_4_kmeans_rarefaction00_a.pdf***. The issue of number of samples and the distribution of their sequencing depths in the context of controlling for uneven sampling effort is explored in far greater detail in an another analysis using sequencing depths observed in biological samples [XXXXXX].

### 8. A filtering step was applied to remove rare taxa from the simulated datasets, which could distort the shape of the communities

@@

Although my previous work does an extensive analysis of the effects of removing rare sequences, I wanted to explore the effect of filtering in the context of the WNWN simulation framework. For each of the filtered and non-filtered OTU tables I calcualted the absolute value of the difference in accuracy between each distance calculation following the normalization procedure (Figure **compare_filter_accuracy.pdf**). With the exception of the Weighted UniFrac distances, each of the distance calculations and normalization procedures were sensitive to the filtering. The rarefaction data tended to be sensitive to filtering at small effect sizes. Distances generated using raw counts, DeSeq variance stabilization, and Upper Quartile Log Fold Change tended to be more sensitive to filtering at larger effect sizes. When using relative abundance data, Bray-Curtis distances were sensitive to filtering at small effect sizes and Unweighted UniFrac distances were sensitive at large effect sizes. These trends appear to be driven by the dependence of the distance calculation on low abundance taxa. More surprising than the effect of filtering on the mean absolute difference in clustering accuracy was the wide variation in accuracies at each effect size. Among the different normalization methods, the accuracies calculated using rarefaction had the narrowest 95% confidence interval for all distance calculations except for calculating Unweighted UniFrac distances. For these distances, relative abundance had the narrowest range at small effect sizes; at larger effect sizes, rarefaction had the narrowest range. Again, these trends appear to be driven by the dependence on low frequency taxa in Unweighted UniFrac, which is dependent on the presence or absence of taxa rather than their abundance. Given my previous work and the large variation caused by removing rare taxa, OTU filtering should not be performed in microbiome analyses.

### 9. No test of effect of transformation methods to account for effects of uneven sample size between treatment groups

@@

### 10. Clustering accuracy was used rather than direct comparisons beta diversity

>>>>

```{r, echo=FALSE, warning = FALSE}
options(tidyverse.quiet = TRUE)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggtext))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(glue))

@@

# Waste not, want not: Revisiting the analysis that called the practice of rarefying microbiome data into question

@@

Since the development of highly parallelized sequencing technologies that enable microbiome researchers to distribute sequences across multiple samples in a single sequencing run, researchers have struggled to produce a consistent number of sequences from each sample in a dataset. It is common to observe more than 10-fold variation in the number of sequences per sample [XXXXX]. Researchers desire strategies to limit uneven sampling effort and need methods to control for he uneveness in their analyses. Of course, uneven sampling is not unique to microbiome research and is a challenge faced by all community ecologists. Common approaches to controlling uneven sampling efforts include use of proportional abundance (i.e., relative abundance), normalization of counts, parameter estimation, and rarefaction.

In 2014 Paul McMurdie and Susan Holmes published "Waste not, want not: why rarefying microbiome data is inadmissible" (WNWN) in PLOS Computational Biology [XXXXX]. This paper has had a significant impact on the approaches that microbiome researchers use to analyze 16S rRNA gene sequence data. According to Google Scholar, this paper has been cited more than 2,400 times as of April 2023. Anecdotely, I have received correspondence from researchers over the past 10 years asking how to address critiques from reviewers who criticize my correspondents' analysis for using rarefaction (e.g., see [this Twitter thread](https://twitter.com/inanna_nalytica/status/1264299305067786242)). I have also received similar comments from reviewers regarding my own work. Most recently, I received the critique for a preprint that I posted in 20XX that analyzes the practice of removing rare taxa from microbiome analyses [XXXXX]. In the process of responding to these reviewers' comments and preparing a manuscript investigating rarefaction and other approaches to control for uneven sequencing effort, I decided to reassess the WNWN study including their definitions, simulations, and analyses.

@@

As I attempted to reproduce the results of WNWN, I noticed that the step that rarefied the data only performed one subsampling of the data. This caused me to re-inspect how McMurdie and Holmes defined "rarefying" in the following quoted text from their paper:

@@

It was unfortunate that McMurdie and Holmes used the term "rarefying" here and throughout their manuscript. The authors were correct to state that the distinction between "rarefying" and "rarefaction" is confusing and leads to their conflation. Adding to the confusion is that the papers cited in the first sentence of this quote (i.e., their references [17]–[19]) either do not use the words "rarefy" or "rarefying" or use them interchangably with "rarefaction". In my experience, subsequent researchers have conflated the results of this study of the effects of rarefying data with rarefaction of data. As an example, Willis (XXXXX) describes problems with rarefaction rather than rarefying data when citing WNWN in her paper proposing alternatives to rarefaction for use with alpha diversity data:

@@

In hindsight, as shown in the quoted text from WNWN, McMurdie and Holmes did emphasize the distinction between rarefying and rarefaction. However, because they seem to have coined a new meaning for rarefying, they added to the confusion by using the generally used verb form of rarefaction. Further confusion comes from the author's admonition in the final sentence that some researchers have failed to repeat the subsampling step. To most scientists, repeating the subsampling step is rarefaction. My preference is to use *subsampling* as the term describing the process they refer to as rarefying. In other words, subsampling is rarefaction, but with a single randomization. To minimize confusion, I will use subsamping in place of rarefying.

I propose the following definition of rarefaction:

@@

This definition aligns well with how rarefaction was originally defined for comparing richness (i.e., the number of taxa in a community) across communities when communities are sampled to different depths [XXXXXXXXX]. With this more general approach to rarefaction, rarefaction can be performed using any alpha or beta diversity metric. This strategy has been widely used by my research group and others and is available in the mothur software package using commands such as `summary.single`, `rarefaction.single`, `dist.shared`. The procedure outlined above could also be used for hypothesis tests of differential abundance; however, thought would need to be given to how to synthesize the results of these tests across a large number of replications.

@@

McMurdie and Holmes analyzed the effect of rarefying and other approaches on clustering accuracy using what they called "Simulation A" in their Figure 2A and elsewhere in their paper. In Simulation A, they investigated the ability to correctly assign simulated microbiome samples to one of two clusters representing two simulated treatment groups. Thankfully, the R code for Simulation A was provided by the authors in the `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` R-flavored markdown file that was published as Protocol S1 in the original paper. I will outline the simulation strategy below and will reference line numbers from their R-flavored markdown file with "L" as a prefix.

For the WNWN cluster analysis, sampling distributions for the two treatment groups were generated using human fecal and ocean data originally take from the GlobalPaterns dataset (L129) [XXXXX]. To generate a fecal and ocean distribution, the authors included any operational taxonomic unit (OTU) that appeared in more than one of the 4 fecal and 3 ocean samples (L60 and L137). The OTUs were sorted by how many of the 7 samples the OTUs were observed in followed by their total abundance across all 7 samples (L139). From this sorted list they identified the identifiers of the first 2000 OTUs (L66). Returning to the 7 samples they selected the data for the corresponding 2000 OTUs and pooled the OTU abundances of the fecal and ocean samples separately to create two sampling distributions (L144, L159-160, L197-198). Next, the fecal and ocean distributions were mixed in 8 different fractions to generate two community types that differend by varying effect sizes (i.e., 1, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5; L170-195, L220); an effect size of 1 generated a null model with no difference between the treatment groups. To simulate the variation in sequencing depth across the 80 samples, they normalized he number of sequences from each of the `r gp_n_samples` samples in the GlobalPatterns dataset so that the median number of sequences (N~L~) for the GlobalPatterns had 1,000, 2,000, 5,000, or 10,000 sequences (L324-325). They then randomly sampled the `r gp_n_samples` normalized sequencing depths to generate 80 sampling depths. From each community type, they simulated 40 samples by sampling to the desired number of sequences (L73, L230-233 and L326-327). Each simulation condition was repeated 5 times (L85). This resulted in 160 simulations (8 effect sizes x 4 median sampling depths x 5 replicates = 160 simulations). Finally, they removed rare and low prevalence OTUs in two steps. First, they removed any OTUs whose total abundance was less than 3 across all 80 samples and that did not appear in at least 3 samples (L368-386). Second, they removed any OTUs that did not have more than 1 sequence in more than 5% of the 80 samples (i.e., 4 samples) and that did not have a total abundance across the 80 samples greater than one half of the number of samples in each community type (i.e., 20) (L523-538, L551).

After generating the OTU counts for the 160 simulated communities, the authors applied several normalization methods, distance calculations, and clustering algorithms to the data. To normalize the OTU data the original analysis either applied no normalization procedure, calculated OTU relative abundances, subsampled the data, performed variance stabilization using the DESeq R package, or performed Upper-Quartile log-fold change normalization using the edgeR R package. In subsampling the data, the authors either included all of the samples or removed samples whose sequencing depth fell below the 5, 10, 15, 20, 25, and 40 percentile across all 80 samples. Subsampled data were then used as input to calculate distances between samples using Bray-Curtis, Euclidean, Unweighted UniFrac, and Weighted Unifrac distances as implemented in the Phyloseq R package, Poisson distance as implemented in the PoiClaClu R package, and top-mean squared difference as implemented in the edgeR R package. Un-normalized data were used to calculate Bray-Curtis, Euclidean, Poisson, and Weighted UniFrac distances. Relative abundance data were used to calculate Bray-Curtis, Unweigthed UniFrac, and Weighted UniFrac distances. DESeq variance stabilization normalized data were used to calculate Bray-Curtis, Euclidean, and Weighted Unifrac distances. The Upper-Quartile log-fold change normalized data were only used to calculate top-mean squared difference distances. The resulting distance matrices were used to cluster the 80 samples into one of two clusters using partitioning around the medoid (PAM), K-means clustering, and hierarchical clustering. Although data for all three methods were presented in the supplementary Protocol S1, only the PAM data are presented in the main manuscript. The accuracy of the clustering assignments was quantified as the fraction of the 80 samples that were assigned to the correct cluster. Because some of the subsampling conditions removed samples those were counted as mis-clustered samples yielding minimum accuracies below 50%.

@@

1. Simulated conditions were only replicated 5 times each
2. The average sizes of the libraries were small by modern standards
3. DESeq-based variance stabilization was used with distance calculation methods that are sensitive to negative values
4. A single subsampling of each dataset was evaluated rather than using rarefaction, which likely resulted in noisier data
5. Results using PAM clustering were not directly compared to those of K-means and hierarchical clustering
6. Subsampling removed the smallest 15% of the samples, which penalized accuracy values by 15 percentage points
7. The distribution of library sizes was not typical of those commonly seen in microbiome analyses
8. A filtering step was applied to remove rare taxa from the simulated datasets
9. No accounting of case when library sizes are confounded with treatment group
10. Clustering accuracy was used rather than direct comparisons of beta diversity
11. There was no consideration of effects of normalization methods on alpha diversity metrics

@@

Each simulated condition was replicated 5 times in WNWN and the paper reports the mean and standard deviation of the replicate clustering accuracies. The relatively small number of replicates accounts for the jerkiness of the lines in the original Figures 4 and 5 (e.g. the Bray-Curtis distances calculated on the DESeqVS normalized data). A better approach would have been to use 100 replicates as this would reduce the dependency of the results on the random number generator's seed. By increasing the number of replicates it was also possible to compare the probability of falsely and correctly clustering samples from the same and different treatment groups together (see points 9-11, below). Because the accuracies are unlikely to be symmetric around a mean at larger accuracy values the median and 95% confidence intervals or intraquartile range should have been reported. To test the effect of increasing the number of replicates, I pulled apart the code in `simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd` into individual R and bash scripts that were executed using a Snakemake workflow with the same Conda environment as above. This was necessary since the number of simulated conditions because of this change increased 20-fold. Such intense data processing was not practical within a single R-flavored markdown document. Again, the observed results were qualitiatively similar to those generated using the single R-flavored markdown file (Figures S4 (**pam_subsample15_fig_4.pdf**) and S5(**pam_subsample_fig_5.pdf**) ). The increased number of replications resulted in smoother lines and allowed me to present empirical 95% confidence intervals. For all analyses in the remainder of this paper, I used 100 randomized replicates per condition.

@@

### 3. DESeq-based variance stabilization was used with distance calculation methods that are sensitive to negative values

Close comparison of the original Figure 4 and my version (**pam_subsample15_fig_4.pdf**) revealed one important difference between the two plots. In the original analysis, the accuracies for the Weighted UniFrac distances at the largest effect size (i.e., 3.5) were 1.0 for median sequencing depths of `r oxford_comma(deseq_wunifrac$n_seqs)`. In my version of the analysis, the values for the same sequencing depths were `r oxford_comma(deseq_wunifrac$accuracy)`, respectively. The 95% confidence interval for these accuracies spaned `r min(deseq_wunifrac$lci)` and `r max(deseq_wunifrac$uci)`. The Bray-Curtis distances were also different by both methods at smaller effect sizes and had wide confidence intervals. Inspection of the DESeq normalized OTU counts revealed that the method resulted in negative values. In fact, rendering the R-flavored markdown files in WNWN's Protocol S1 generated a warning message when passing the DESeq normalized counts to the Bray-Curtis calculator, which said, "results may be meaningless because data have negative entries in method 'bray'". Although the weighted UniFrac calculator function did not generate a similar warning message, negative values would also result in similarly meaningless distances. Both are due to the fact that the calculators sum the counts of each OTU in both samples being compared. In contrast, a Euclidean distance does not use a similar sum, but sums the square of the difference between the OTU abundance in each sample.

To assess the prevalence of negative counts in the simulated data, I quantified the fraction of negative values in the OTU matrix from each simulation and counted the number of simulations where the normalized OTU table had at least one negative value (Figure **deseq_negative_value.pdf**). In general the fraction of negative OTU counts increased with effect size, but decreased with sequencing effort. The fraction of simulations with at least onee negative value increased with effect size and sequencing effort. The high frequency of negative OTU counts resulted in highly variable Bray-Curtis and weighted UniFrac values. It is likely that because the WNWN analysis only used 5 replicates that the large variation in accuracies at high effect sizes was missed initially. For the rest of this reanalysis study, I will only report results using the DESeq-based variance stabilization normalization with the Euclidean distance.

### 4. A single subsampling of each dataset was evaluated rather than using rarefaction, which likely resulted in noisier data

@@

### 6. Subsampling removed the smallest 15% of the samples, which penalized accuracy values by 15 percentage points

In WNWN, the authors quantified the tradeoff between sampling effort, the number of samples removed below the threshold, and clustering accuracy (original Figure 5, my Figure S5). Although the optimal sampling effort varied by distance metric, normalization method, and sampling effort, they removed samples whose number of sequences was less than the 15th percentile (L404-419). They acknowledged that this screening step, which was only used with subsampling, would decrease clustering accuracy putting it at a relative disadvantage to the other methods (page 5, column 1, last paragraph). Therefore, it was not surprising that the peak clustering accuracy for their subsampled data was at 85% (Figure S2). Because the true result would not known *a priori* in microbiome studies, it would be impossible for researchers to conduct a sensitivity analysis comparing the tradeoffs between sequencing depth, sample number, and clustering accuracy to select a sampling depth. [**Although the authors claim that "Rarefying counts requires an arbitrary selection of a library size minimum that affects downstream inference" (page 8, column 1, point 3), in acutal microbiome studies the selection of a sampling depth is not as arbitrary as the authors claim. Rather, to avoid p-hacking, researchers pick a set of criteria where they will include or exclude samples prior to testing their data.**] The differences in clustering accuracy between subsampling and rarefaction and using PAM and K-means clustering indicated that it was necessary to reassess the tradeoff between the library size minimum quantile and clustering accuracy. When using rarefaction, K-means clustering, and only considering conditions with 2,000 or more sequences, there was not a condition where setting a higher threshold resulted in a better accuracy than using all of the samples (**kmeans_rarefaction_fig_5.pdf**). These results showed that for modern sequencing depths, using the full datasets with rarefaction and K-means clustering resulted in accuracies that were typically better than those observed when removing the smallest 15% of the samples from each simulated dataset. When the original Figure 4 was recast with these approaches, Rarefaction performed at least as well as any of the other tranformations with each distance calculation, except for with the Poisson distance (Figure **kmeans_rarefaction15_fig_4.pdf**). It is worth noting that at larger effect sizes, K-means clustering did not perform as well for some combinations of normalization methods and distance calculations (compare **kmeans_rarefaction15_fig_4.pdf** and **pam_subsample00_fig_4.pdf**); however, those combinations that performed worse by K-means were not as good as rarefaction or subsampling by either clustering method.

@@

### 7. The distribution of library sizes was not typical of those commonly seen in microbiome analyses

As described above, the sequencing depths used in the `r gp_n_samples` GlobalPatterns datasets were used as the distribution to create sequencing depths for the 80 samples that were generated in each simulation. The GlobalPatterns datasets had a mean of `r gp_mean` sequences and a median of `r gp_median` sequences per dataset. The datasets ranged in sequencing depth between `r gp_min` and `r gp_max` sequences for a `r gp_max/gp_min`-fold difference. Rather than representing a typically observed distribution of sequencing depths that would be skewed right, the sampling distribution was normally distributed (Shapiro-Wilk test of normality, P=`r shapiro.test(gp_counts$n_seqs)$p.value`) (Figure ***distribution_shape.pdf***). From these simulations it is unclear how sensitive the various normalizations and distance calculations are to a skewed distribution. A second limitation of this sampling distribution is that it only contained `r gp_n_samples` unique sampling depths such that each sampling depth would have been re-used an average of `r 80/gp_n_samples` times in each simulation. Yet, it is unlikely for a real sequence collection to have duplicate sequencing depths.

I created a new set of simulations that would focus on solely on the effect of the shape of the distribution on the results to reassess the WNWN results in the context of a more typical distribution of sample sizes. I created a simple sequencing depth distribution where there were 80 depths logarithmically distributed between the minimum and maximum sequencing depths of the GlobalPatterns dataset (Figure ***distribution_shape.pdf***). The median of this distribution was `r log_median` and the mean was `r log_mean`. When I regenerated the original Figures 4 and 5 using the log-distributed sequencing effort distribution, the differences in normalization methods were more apparent (Figure ***fig_4_kmeans_rarefaction00_log.pdf***). For each of the distance calculators, rarefying the data to the size of the smallest dataset yielded accuracies that were at least as good as the other methods across effect sizes and median sequencing depths. The difference was most pronounced at smaller effect sizes and sequencing depths. When comparing the performance of rarefaction across distance calculators for different effect sizes, sequencing depths and size of smallest sample (***fig_5_kmeans_rarefaction_log.pdf***), the accuracies I observed using the log-distributed sample sizes was at least as good as those obtained using the GlobalPatterns-based distribution ***fig_4_kmeans_rarefaction00_a.pdf***. The issue of number of samples and the distribution of their sequencing depths in the context of controlling for uneven sampling effort is explored in far greater detail in an another analysis using sequencing depths observed in biological samples [XXXXXX].

### 8. A filtering step was applied to remove rare taxa from the simulated datasets

@@

Although my previous work does an extensive analysis of the effects of removing rare sequences, I wanted to explore the effect of filtering in the context of the WNWN simulation framework. For each of the filtered and non-filtered OTU tables I calcualted the absolute value of the difference in accuracy between each distance calculation following the normalization procedure (Figure **compare_filter_accuracy.pdf**). With the exception of the Weighted UniFrac distances, each of the distance calculations and normalization procedures were sensitive to the filtering. The rarefaction data tended to be sensitive to filtering at small effect sizes. Distances generated using raw counts, DESeq variance stabilization, and Upper Quartile Log Fold Change tended to be more sensitive to filtering at larger effect sizes. When using relative abundance data, Bray-Curtis distances were sensitive to filtering at small effect sizes and Unweighted UniFrac distances were sensitive at large effect sizes. These trends appear to be driven by the dependence of the distance calculation on low abundance taxa. More surprising than the effect of filtering on the mean absolute difference in clustering accuracy was the wide variation in accuracies at each effect size. Among the different normalization methods, the accuracies calculated using rarefaction had the narrowest 95% confidence interval for all distance calculations except for calculating Unweighted UniFrac distances. For these distances, relative abundance had the narrowest range at small effect sizes; at larger effect sizes, rarefaction had the narrowest range. Again, these trends appear to be driven by the dependence on low frequency taxa in Unweighted UniFrac, which is dependent on the presence or absence of taxa rather than their abundance. Given my previous work and the large variation caused by removing rare taxa, OTU filtering should not be performed in microbiome analyses.

### 9. No accounting of case when library sizes are confounded with treatment group

@@

### 10. Clustering accuracy was used rather than direct comparisons of beta diversity
